---
title: "MLM, exerc 12: repeated measurements Sitka spruce data"
author: "Gerrit Gort"
date: July 2, 2021
output: pdf_document
fontsize: 12pt
geometry: margin=0.75in
---

```{r options, include=FALSE, warning=FALSE}
options(digits=7)
options(show.signif.stars=F)
library(MASS)
library(knitr)
opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=TRUE, warning=FALSE, message=FALSE)
knitr::opts_knit$set(root.dir="C:/Users/gort001/OneDrive - Wageningen University & Research/Onderwijs/PhD-courses/AdvancedStatisticsModuleMLM/course2021_MLM/data")
```

# Read data

```{r}
Sitka <- read.csv("exercise12.csv")
Sitka$tree <- factor(Sitka$tree)
Sitka$treat <- factor(Sitka$treat)
head(Sitka)
```


# Plot of size versus time per tree split by treatment

```{r, fig.width=9, fig.height=6, strip.white=TRUE}
library(nlme)
Sitka.grp <- groupedData(size ~ time | tree, outer= ~treat, data=Sitka)
# nice plot of size vs time
plot(Sitka.grp, outer= ~treat, key=FALSE, aspect=0.85, layout=c(2,1), 
     xlab=list(fontsize=20), ylab=list(fontsize=20), scales=list(cex=1.25))
```

# Fitting different covariance structures

## 1: Compound symmetry: constant correlation among observations from same tree
```{r}
(lme.tree1 <- gls(size ~ treat + time + treat:time, correlation =  corCompSymm(form = ~ 1 | tree), data=Sitka) ) 
summary(lme.tree1)
logLik(lme.tree1)
AIC(lme.tree1)

# Different specification of same model
(lme.tree1.1 <- lme(size ~ treat + time + treat:time, random= ~ 1  | tree, data=Sitka) ) 
summary(lme.tree1.1)
logLik(lme.tree1.1)
AIC(lme.tree1.1)
```
## 2: AR(1): autocorrelation order 1
```{r}
(lme.tree2 <- gls(size ~ treat + time + treat:time, 
                  correlation = corAR1(form = ~ 1 | tree), data=Sitka) )
summary(lme.tree2)
logLik(lme.tree2)
AIC(lme.tree2)
```
## 3: Unstructured covariance matrix

a. unstructured covariance matrix with heterogeneous variances

b. unstructured covariance matrix with constant variance

```{r}
(lme.tree3a <- gls(size ~ treat + time + treat:time, 
                  correlation =  corSymm(form = ~ 1 | tree), 
                  weights=varIdent(form= ~ 1 | time), data=Sitka) ) 
summary(lme.tree3a)
logLik(lme.tree3a)
AIC(lme.tree3a)

(lme.tree3b <- lme(size ~ treat + time + treat:time, 
                  random = list(tree = pdSymm(~ 1)), data=Sitka) ) 
summary(lme.tree3b)
logLik(lme.tree3b)
AIC(lme.tree3b)
```
## 4: random coefficient model
```{r}
lme.tree4 <- lme(size ~ treat + time + treat:time, random= ~ 1 + time | tree, data=Sitka)
summary(lme.tree4)
logLik(lme.tree4)
AIC(lme.tree4)
```

Comparison of AIC's:

- model 1 (compound symmetry): `r round(AIC(lme.tree1),1)`

- model 2 (autoregressive order 1): `r round(AIC(lme.tree2),1)`

- model 3a (unstructured, heterogeneous variance): `r round(AIC(lme.tree3a),1)`

- model 3b (unstructured, common variance): `r round(AIC(lme.tree3b),1)`

- model 4 (random coefficients): `r round(AIC(lme.tree4),1)`

Notice that the unstructured covariance matrix with heterogeneous variances gives by far the lowest AIC, meaning best fitting model. This is a quite costly structure with a lot parameters, though.

# Compare F-values of fixed effects
```{r}
anova(lme.tree1)
anova(lme.tree2)
anova(lme.tree3a)
anova(lme.tree3b)
anova(lme.tree4)
```

Notice that hypothesis test results for fixed effects do depend on the chosen covariance structure.

# Extract variance covariance matrices
```{r}
round(getVarCov(lme.tree1),2)
round(getVarCov(lme.tree2),2)
round(getVarCov(lme.tree3a),2)
round(getVarCov(lme.tree3b),2)

getVarCov(lme.tree4, type="marginal")

getVarCov(lme.tree4, type="random.effects")
# This is the covariance matrix of intercepts - slopes
```


# To get correlation matrix do e.g.
```{r}
covmat3a <- getVarCov(lme.tree3a)
std3a <- sqrt(diag(covmat3a))
corrmat3a <- diag(1/std3a) %*% covmat3a %*% diag(1/std3a)
round(corrmat3a,2)

# Do the same for the other 4 covariance matrices...
```

Notice the rather unexpected pattern for the unstructured correlation matrix: correlation decreases until the third timepoint, but then increases again.

This may be related to a failing systematic part of the model. In the starting plot in this report clearly curved relationships can be seen. You could try to add a quadratic effect of time and next refit the models with the different covariance structures.

Try yourself...